--- 
title: "Notes on Biostatistics"
author: "Phùng Khánh Lâm"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: lampk/notebook_Biostatistics
description: "My notes on Biostatistics"
---

# Preface {-}


<!--chapter:end:index.Rmd-->

# Bookdown {-}

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

<!--chapter:end:00-bookdown.Rmd-->

# New {#new}

* Bianchi, M. T., Alexander, B. M., & Cash, S. S. (2009). Incorporating Uncertainty Into Medical Decision Making: An Approach to Unexpected Test Results. Medical Decision Making, 29(1), 116–124. https://doi.org/10.1177/0272989X08323620

* Nobles AL, Leas EC, Althouse BM, et al. Requests for Diagnoses of Sexually Transmitted Diseases on a Social Media Platform. JAMA. 2019;322(17):1712–1713. doi:https://doi.org/10.1001/jama.2019.14390

* Steve Ruberg (2019) No. 13: Unconsciously Biased and Consciously Unbiased. https://analytixthinking.blog/2019/11/07/no-13-unconsciously-biased-and-consciously-unbiased/

* renv: Project Environments for R. https://blog.rstudio.com/2019/11/06/renv-project-environments-for-r/

* The frontier of simulation-based inference. https://deepai.org/publication/the-frontier-of-simulation-based-inference

* Non-randomized studies using causal-modelling may give different answers than RCTs: a meta-epidemiological study. DOI: https://doi.org/10.1016/j.jclinepi.2019.10.012. 

* Four mistakes to avoid when talking with patients about risk. https://www.aafp.org/journals/fpm/blogs/inpractice/entry/risk_communication_mistakes.html#.XDYKYKwH6CA.twitter

* 5 mistakes of communicating risk. https://www.youtube.com/watch?v=jAY3H92MXGU

*  @BenVanCalster's keynote on Machine learning: calm down!. https://twitter.com/Georg__Heinze/status/1192796427619065857

* What logic gives us the authority to say something about an individual by observing other individuals, however similar? https://twitter.com/yudapearl/status/1192768005316341760. http://ftp.cs.ucla.edu/pub/stat_ser/r375-reprint.pdf

* Just released today, and it is shameful: https://ncd.gov/sites/default/files/NCD_Quality_Adjusted_Life_Report_508.pdf.  Recommendation against QALYs and #healtheconomics  to support resource allocation decisions. https://twitter.com/dollendorf/status/1192181668985102338

* stats might kill personalized medicine. https://twitter.com/KertViele/status/1192194206049226754

* RStudio Package Manager. https://blog.rstudio.com/2019/11/07/package-manager-v1-1-no-interruptions/



<!--chapter:end:00-new.Rmd-->

# Initial data analysis {#ida}


<!--chapter:end:01-ida.Rmd-->

# Data cleaning {#DataCleaning}

<!--chapter:end:02-datacleaning.Rmd-->

# Descriptive analysis {#DescriptiveAnalysis}

<!--chapter:end:03-DescriptiveAnalysis.Rmd-->

# Graphics {#graphics}

<!--chapter:end:04-graphics.Rmd-->

# Causal Inference {#CausalInference}

## Reading list

* Daniel Westreich (2019) Epidemiology by Design: A Causal Approach to the Health Sciences. Oxford University Press. https://global.oup.com/academic/product/epidemiology-by-design-9780190665760?cc=vn&lang=en&

<!--chapter:end:05-CausalInference.Rmd-->

# Study design {#StudyDesign}

<!--chapter:end:06-design.Rmd-->

# Hypothesis Testing {#HypothesisTesting}

<!--chapter:end:07-HypothesisTesting.Rmd-->

# P value {#Pvalue}

<!--chapter:end:08-p.Rmd-->

# Bayesian data analysis {#Bayesian}

<!--chapter:end:09-Bayesian.Rmd-->

# Prediction models {#PredictionModels}

<!--chapter:end:10-PredictionModels.Rmd-->

# Medical decision making {#mdm}

## Reading list

### Risk communication

* Sisk BA, Baker JN (2018) Microethics of Communication—Hidden Roles of Bias and Heuristics in the Words We Choose. JAMA Pediatr;172(12):1115–1116. doi:https://doi.org/10.1001/jamapediatrics.2018.3111.
* Steinhardt Joe (2019) The role of numeric and statistical content on risk perception in infographics about road safety, Journal of Risk Research, DOI:10.1080/13669877.2019.1596147
* Kunneman M, Stiggelbout AM, Pieterse AH (2019) Do clinicians convey what they intend? Lay interpretation of verbal risk labels used in decision encounters. Patient Educ Couns. 2019 Aug 26. pii: S0738-3991(19)30377-5. doi: 10.1016/j.pec.2019.08.035.
* NICE guidance on risk communication: https://twitter.com/BerksMaternity/status/1171542152842817536/photo/1. [Figure](figures/mdm/nicenice_riskcommunication.jpeg)
* Okan Y, Smith SG, Bruine de Bruin W (2019) How is cervical cancer screening information communicated in UK websites? Cross-sectional analysis of content and quantitative presentation formatsBMJ Open 2019;9:e029551. doi: 10.1136/bmjopen-2019-029551.
* Yen, Renata W, Barr, Paul J, Cochran, Nan, Aarts, Johanna W, Légaré, France, Reed, Malcolm, O'Mallley, A James, Scalia, Peter, Guérard, Genevieve Painchaud, Backer, Grant, Reilly, Clifford, Elwyn, Glyn and Durand, Marie-Anne (2019) Medical students’ knowledge and attitudes towards shared decision-making: results from a multinational cross-sectional survey. Medical Decision Making. http://sro.sussex.ac.uk/id/eprint/85728/
* 


<!--chapter:end:11-MedicalDecisionMaking.Rmd-->

# Intensive Longitudinal data analysis {#ild}

## Reading list

### Papers:

* T. Asparouhov, E. L. Hamaker, and B. Muthén. “Dynamic Structural Equation Models”. In: _Structural Equation Modeling:
A Multidisciplinary Journal_ 25.3 (Dec. 2017), pp. 359-388. DOI: 10.1080/10705511.2017.1406803. <URL:
https://doi.org/10.1080/10705511.2017.1406803>.

* T. Asparouhov and B. Muthén. “Comparison of Models for the Analysis of Intensive Longitudinal Data”. In: _Structural
Equation Modeling: A Multidisciplinary Journal_ (Jul. 2019), pp. 1-23. DOI: 10.1080/10705511.2019.1626733. <URL:
https://doi.org/10.1080/10705511.2019.1626733>.

* E. M. Chi and G. C. Reinsel. “Models for Longitudinal Data with Random Effects and AR(1) Errors”. In: _Journal of the
American Statistical Association_ 84.406 (Jun. 1989), pp. 452-459. DOI: 10.1080/01621459.1989.10478790. <URL:
https://doi.org/10.1080/01621459.1989.10478790>.

* L. M. Collins. “Analysis of Longitudinal Data: The Integration of Theoretical Model, Temporal Design, and Statistical
Model”. In: _Annual Review of Psychology_ 57.1 (Jan. 2006), pp. 505-528. DOI: 10.1146/annurev.psych.57.102904.190146.
<URL: https://doi.org/10.1146/annurev.psych.57.102904.190146>.

* S. C. Duncan, T. E. Duncan, and H. Hops. “Analysis of longitudinal data within accelerated longitudinal designs.” In:
_Psychological Methods_ 1.3 (1996), pp. 236-248. DOI: 10.1037/1082-989x.1.3.236. <URL:
https://doi.org/10.1037/1082-989x.1.3.236>.

* T. E. Duncan and S. C. Duncan. “An introduction to latent growth curve modeling”. In: _Behavior Therapy_ 35.2 (2004),
pp. 333-363. DOI: 10.1016/s0005-7894(04)80042-x. <URL: https://doi.org/10.1016/s0005-7894(04)80042-x>.

* E. L. Hamaker and M. Wichers. “No Time Like the Present”. In: _Current Directions in Psychological Science_ 26.1
(Feb. 2017), pp. 10-15. DOI: 10.1177/0963721416666518. <URL: https://doi.org/10.1177/0963721416666518>.

* E. L. Hamaker, T. Asparouhov, A. Brose, et al. “At the Frontiers of Modeling Intensive Longitudinal Data: Dynamic
Structural Equation Models for the Affective Measurements from the COGITO Study”. In: _Multivariate Behavioral Research_
53.6 (Apr. 2018), pp. 820-841. DOI: 10.1080/00273171.2018.1446819. <URL: https://doi.org/10.1080/00273171.2018.1446819>.

* N. C. Jacobson, S. Chow, and M. G. Newman. “The Differential Time-Varying Effect Model (DTVEM): A tool for diagnosing
and modeling time lags in intensive longitudinal data”. In: _Behavior Research Methods_ 51.1 (Aug. 2018), pp. 295-315.
DOI: 10.3758/s13428-018-1101-0. <URL: https://doi.org/10.3758/s13428-018-1101-0>.

* S. Jahng and P. K. Wood. “Multilevel Models for Intensive Longitudinal Data with Heterogeneous Autoregressive Errors:
The Effect of Misspecification and Correction with Cholesky Transformation”. In: _Frontiers in Psychology_ 8 (Feb. 2017).
DOI: 10.3389/fpsyg.2017.00262. <URL: https://doi.org/10.3389/fpsyg.2017.00262>.

* M. Karas, J. Bai, M. Strączkiewicz, et al. “Accelerometry Data in Health Research: Challenges and Opportunities”.
In: _Statistics in Biosciences_ 11.2 (Jan. 2019), pp. 210-237. DOI: 10.1007/s12561-018-9227-2. <URL:
https://doi.org/10.1007/s12561-018-9227-2>.

* S. T. Lanza, S. A. Vasilenko, and M. A. Russell. “Time-varying effect modeling to address new questions in
behavioral research: Examples in marijuana use.” In: _Psychology of Addictive Behaviors_ 30.8 (Dec. 2016), pp. 939-954.
DOI: 10.1037/adb0000208. <URL: https://doi.org/10.1037/adb0000208>.

* Y. Miyazaki and S. W. Raudenbush. “Tests for linkage of multiple cohorts in an accelerated longitudinal design.” In:
_Psychological Methods_ 5.1 (2000), pp. 44-63. DOI: 10.1037/1082-989x.5.1.44. <URL:
https://doi.org/10.1037/1082-989x.5.1.44>.

* S. W. RAUDENBUSH and W. CHAN. “Growth Curve Analysis in Accelerated Longitudinal Designs”. In: _Journal of Research
in Crime and Delinquency_ 29.4 (Nov. 1992), pp. 387-411. DOI: 10.1177/0022427892029004001. <URL:
https://doi.org/10.1177/0022427892029004001>.

* J. Windt, C. L. Ardern, T. J. Gabbett, et al. “Getting the most out of intensive longitudinal data: a methodological
review of workload\textendashinjury studies”. In: _BMJ Open_ 8.10 (Oct. 2018), p. e022626. DOI:
10.1136/bmjopen-2018-022626. <URL: https://doi.org/10.1136/bmjopen-2018-022626>.

* S. Yang, J. A. Cranford, R. Li, et al. “A time-varying effect model for studying gender differences in health
behavior”. In: _Statistical Methods in Medical Research_ 26.6 (Oct. 2015), pp. 2812-2820. DOI: 10.1177/0962280215610608.
<URL: https://doi.org/10.1177/0962280215610608>.

### Books

* Walls T, Schafer J (eds.) (2006). _Models for intensive longitudinal data_. Oxford University Press.

* K. van Montfort, J. H. Oud, and M. C. Voelkle, ed. _Continuous time modeling in the behavioural and related
sciences_. Springer, 2019.

* O. Ryan, R. Kuiper, and E. Hamaker. “A continuous time approach to intensive longitudinal data: What, Why and How?”
In: _Continuous time modeling in the behavioural and related sciences_. Springer, 2019. Chap. 2.

### Software

* L. Ou, M. D. Hunter, and S. Chow. “What's for dynr: A package for linear and nonlinear dynamic modeling in R”. In:
_The R Journal_ (2019), pp. 1-20. DOI: 10.32614/RJ-2019-012.

### Presentations

* B. Muthen, T. Asparouhov, and E. Hamaker. _Which methods do we need for intensive longitudinal data_. 2016.

* N. Schuurman. _Analyzing Intensive Longitudinal Data_. 2018.

<!--chapter:end:12-ild.Rmd-->

# Softwares {#softwares}

## R

* library(printr). https://twitter.com/robinson_es/status/1192875040393629696
* Javier Luraschi, Kevin Kuo, Edgar Ruiz (2019) Mastering Apache Spark with R. https://therinspark.com/?utm_content=buffer62cba&utm_medium=social&utm_source=twitter&utm_campaign=buffer

<!--chapter:end:13-softwares.Rmd-->

`r if (knitr:::is_html_output()) '
# References {-}
'`

<!--chapter:end:99-references.Rmd-->

